Towards Semantic Patches

### 2010-06-18 papers/MIPS2010/abstract.tex
Reactive les commandes de commentaire.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@105 403e655e-980a-47bd-8c60-1ee5348bb6b7
fix intro, add related works (a tailler)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@106 403e655e-980a-47bd-8c60-1ee5348bb6b7

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\usepackage[all,cmtip]{xy}
%\usepackage[protrusion=true,expansion=true]{microtype}

@@

% Commandes à désactiver avant la soumission.
% Remarques sur la structure.
%\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
% Remarques sur le contenu.
%\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

\newcommand{\remplan}[1]{}
\newcommand{\remtext}[1]{}

@@

\author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
\date{}

@@

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.
% \textsf{Coq} or \textsf{Matita}.

\remtext{Fri Jun 18, 2010 9:55. AM Ici, on pourrait rajouter un
  paragraphe pour dire qu'il existe des travaux sur la formalisation
  de metatheories dans les assistants de preuve, que ce soit en TWELF
  ou Coq. Puis, on explicite notre but général qui est de construire
  un système/framework qui remplace la chaîne d'outils traditionnelle
  en important/adaptant la technologie issue de la formalisation de
  metatheories.}

@@

We propose to discuss a small part of these questions, namely the
enhancement and adaptation of version control paradigms to the
management of mathematical repositories, to witness with more
precision the \emph{impact of changes}. By studying the dependencies
between objects in a proof, we can get a much finely grained vision of
what a future evolution would impact. In proof assistants based on the
propositions-as-types paradigm, Type Theory offers a powerful tool to
witness the effect of changes finely, \emph{i.e.} types. Eventually,
we hope to be capable of expressing evolutions on formal proofs, check
them efficiently and incrementally, and manage distributed
repositories of mathematics with it, guaranteeing \emph{by typing} the
stepwise global consistency of the repository.

\remtext{Fri Jun 18, 2010 10:06 AM.  Le paragraphe précédent n'est pas
  assez précis je pense et en même temps en dit trop. À la place, je
  propose d'énoncer succintement les contributions de ce papier puis 
  le plan. Ce serait quelque chose comme:
  1. A description language for semantic repositories.
  2. Theory development through semantic patches. 
  3. Related work
  4. Further work
}

\remplan{Fri Jun 18, 2010 10:08 AM. Du coup, je restructure un peu la suite.}

\section{A kernel language to describe semantic repositories}

@@

\section{Development through semantic patches}

@@

\section{Related work}

\remtext{Dolphin, Twelf, etc}

\remtext{Semantic patches à la coccinnelle}

\section{Further work}

>>>>

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\usepackage[all,cmtip]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}

@@

% Commandes à désactiver avant la soumission.
% Remarques sur la structure.
\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
% Remarques sur le contenu.
\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

%\newcommand{\remplan}[1]{}
%\newcommand{\remtext}[1]{}

@@

\author{Matthias Puech, Yann Régis-Gianas}
\date{}

@@

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.

In fact, although dedicated tools exist to formalize the description
of languages and their metatheory (e.g. \textsf{Twelf},
{\cite{pfenning1999system}}), and substantial formalizations have been
undertaken \remtext{TODO citer, mais quoi?}, we still use legacy tools
based on text representation to manage our developments. The general
goal exposed here is to replace this tool chain and make it
language-aware, both on a syntactic side (AST) and on a semantic side
(typing).

@@

We propose to discuss a small part of these questions, namely the
enhancement and adaptation of version control paradigms to the
management of mathematical repositories, to witness with more
precision the \emph{impact of changes}. In proof assistants based on the
propositions-as-types paradigm, Type Theory offers a powerful tool to
witness the effect of changes finely: types.

We will describe here some of the possible directions to develop a
tool to analyze the impact of change through types. It involves at its
core a typed description language for repositories -- described in
\S\ref{kernel} -- driven by a user interface that we sketch in
\S\ref{devel}. In the first iteration of this project, we focus on a
static, or data-driven model for repositories; in \S\ref{rw}, we
quickly review related approaches, and hint a possible generalization
in \S\ref{fw}.

\section{A kernel language to describe semantic repositories}
\label{kernel}

@@

\section{Development through semantic patches}
\label{devel}

@@

\section{Related work}
\label{rw}

The \textsf{Twelf} project (\cite{pfenning1999system}) is an
implementation of the Logical Framework (LF,
\cite{harper1993framework}). It was used in \cite{anderson1993program}
to devise transformations of proofs in order to extract efficient
programs.

The problems of managing a formal mathematical library have been dealt
with in various proof assistant and mathematical repositories. The
HELM project (\cite{asperti2000content}, \cite{asperti2006content})
was an attempt to create a large library of mathematics, importing
\textsf{Coq}'s developments into a searchable and browsable database.
Most ideas from this project were imported into the \textsf{Matita}
proof assistants (\cite{asperti2007hop}), especially a mechanism of
\emph{invalidation and regeneration} to ensure the global consistency
of its library w.r.t changes, with granularity the whole definitions
or proofs and their dependencies. The MBase project
(\cite{kohlhase2001mbase}) attempts at creating a web-based,
distributed mathematical knowledge database putting forward the idea
of \emph{development graph} (\cite{hutter2000management},
\cite{autexier2000towards}) to manage changes in the database,
allowing semantic-based retrieval and object-level dependency
management.

This idea, generalized over structured, semi-formal documents gave
birth to \texttt{\it locutor} (\cite{muller2008fine}), a fine-grained
extension of the \texttt{svn} version control system for XML
documents, embedding ontology-driven, user-defined semantic knowledge
which allows to go across the filesystem border. It embeds a
\emph{diff} algorithm, operating on the source text modulo some
equality theory to quotient the syntax. On the same line of work, we
should mention the \emph{Coccinelle} tool
(\cite{padioleau2008documenting}). It is an evolution over textual
patches, specialized on the C language, allowing more flexibility in
the matching process, and was developed to deal with the problem of
\emph{collateral evolutions} in the Linux kernel. It embeds a
declarative language for matching and transforming C source code,
operating on text modulo defined isomorphisms.

Our approach to the ``impact of changes'' problem seems novel on
several aspects: first, it applies uniformly on proofs and programming
languages by virtue of the Curry-Howard isomorphism, and because we
operate at the AST level. Secondly, by taking \emph{types} as a
witnesses for the evolution of a development, we refine the usual,
dependency-based approach for a finer granularity.

\section{Further work}
\label{fw}
